---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PhyloDecR

<!-- badges: start -->
<!-- badges: end -->

The goal of PhyloDecR is to check sets of taxon sets for phylogenetic decisiveness in R.

## Installation

You can install the released version of PhyloDecR from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("PhyloDecR")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pottNeJa/PhyloDecR")
```
## To do

* Check MA for the three conditions (number of quadruples, triples & tuples) &  add the corresponding lemma or theorem
* Test all used taxa sets of the MA
* Optimize algorithm output
* Full description of all the functions
* Test with taxa sets from Mareike
* Compare algorithm with Mareikes Mathematica Code

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(PhyloDecR)

## basic example code
data(exampleData)
head(exampleData)

```

### Step 1: Create Input

```{r step1}
test1 = createInput(fn="../../2103_FischerPaper/Beispiele/example_Fischer_1.txt",sepSym = ",")
test2 = createInput(fn="../../2103_FischerPaper/Beispiele/example_Fischer_2.txt",sepSym = ",")
test3 = createInput(fn="../../2103_FischerPaper/Beispiele/example_8_12_Decisive.txt", sepSym = "_")

test1$input_raw
test1$input_quadruples
test1$input_ordered
test1$taxa

head(test1$data)
# knitr::kable(test1$data,caption = "All possible quadruples with the given taxa and their status (given as input or to be resolved)")

```

### Step 2: Do some initial tests
The second input, *test2*, contains data for 11 taxa lists of 6 different taxa. 

First, we want to check if this input data is okay for the algorithm. There are three conditions (proof see master thesis):

1) Input Quadruple size has to be large enough ($n-1 \choose 4$) --> to check! could also be ($n-1 \choose 3$), maybe just a typo! See Theorem 5 / Conjecture of my Master Thesis
2) All triples must to be in the input data! See Lemma 1 of my Master Thesis.
3) All tuples must be sufficiently available! --> See Theorem 6 of my Master Thesis.

```{r checkInput}
test1_checks = initialCheck(data = test1$data)
test2_checks = initialCheck(data = test2$data)
test3_checks = initialCheck(data = test3$data)

```

Okay, my example data passes all initial tests. This means, that I can run the algorithm created by Mareike Fischer. 

```{r algorithm}
test1_alg<-runAlgorithm(data = test1$data)
test2_alg<-runAlgorithm(data = test2$data)
test3_alg<-runAlgorithm(data = test3$data)
```

