---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PhyloDecR

<!-- badges: start -->
<!-- badges: end -->

The goal of PhyloDecR is to check sets of taxon sets for phylogenetic decisiveness in R.

## Installation

You can install the released version of PhyloDecR from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("PhyloDecR")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pottNeJa/PhyloDecR")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(PhyloDecR)

## basic example code
data(exampleData)
head(exampleData)

```

Okay, we have an example data set with 10 quadruples and n=6 taxa. 

First, we want to check if this input data is okay for the algorithm. There are three conditions (proof see master thesis):

1) Input Quadruple size has to be large enough (> $$ {N\choose 4}$$)
2) All triples must to be in the input data
3) All tuples must be sufficiently available

```{r createInput}
test1<-createInput(fn="_archive/quadruple_check2.txt",sepSym = "_")
test1<-createInput(fn="../../2103_FischerPaper/Beispiele/example_8_12_Decisive.txt",sepSym = "_")

test1$input
test1$taxa
test1$comments

knitr::kable(test1$data,caption = "All possible quadruples with the given n and their status (given as input or to be resolved)")
```

Okay, my example data passes all initial tests. This means, that I can run the algorithm created by Mareike Fischer. 

```{r algorithm, eval=F}
test2<-runAlgorithm(data = test1$data)
```

